---
title: "Problem Set 5"
output: html_document
date: "2023-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r echo=FALSE}
library(tidyverse)
```
## Load and Check Data
### **Question 1**
```{r echo=FALSE}
gapminder <- read.delim ("gapminder.csv.bz2")
```
**gapminder <- read.delim ("gapminder.csv.bz2")**

### **Question 2**
```{r echo=FALSE}
dim (gapminder)
```
### **Question 3**
```{r echo=FALSE}
gapminder [sample (1:nrow (gapminder), 3),]
```
The looks perfect, no issues in sight

## Descriptive Statistics
### **Question 1**
```{r echo=FALSE}
length (unique (gapminder$iso3))
length (unique (gapminder$iso2))
length (unique (gapminder$name))
```
There are 253 countries in variable iso3, 249 countries in variable iso2, and 250 in variable name

### **Question 2**
**A.**
```{r echo=FALSE}
gapminder %>%
  group_by (iso2) %>%
  summarise (n = n_distinct (name)) %>%
  arrange (desc (n))
```
There are no iso2 codes that correspond to more than one country.

**B.**
```{r echo=FALSE}
gapminder %>%
  group_by (name)%>%
  summarise (n = n_distinct (iso3)) %>%
  arrange (desc (n))
```
There are no iso3 codes that correspond to more than one country.

### **Question 3**
```{r echo=FALSE}
max (gapminder$time, na.rm = TRUE)
min (gapminder$time, na.rm = TRUE)
```
The maximum is 2019 and the minimum is 1960

## CO2 Emissions
### **Question 1**
```{r echo=FALSE}
gapminder %>%
  group_by (time) %>%
  summarise (val1 = sum (is.na (co2)), val2 = sum (is.na (co2_PC))) %>%
  arrange (desc (val1))
```
The years 2017, 2018, and 2019 have the most missing data for CO2 and co2_PC

### **Question 2**
```{r echo=FALSE}
gapminder %>% 
  filter (name==c("United States of America", "China", "India", "Uruguay", "France")) %>%
  ggplot (aes(x=time, y=co2, color=name)) + geom_point () + geom_line () + 
  xlab ("Year") + ylab ("CO2 Emissions") + guides (color=guide_legend (title="Countries"))
```

Most countries had a very steady CO2 emissions over time except for China, which increased significantly shortly after the year 2000

### **Question 3**
```{r echo=FALSE}
gapminder %>% 
  filter (name == c("United States of America", "China", "India", "Uruguay", "France")) %>%
  ggplot (aes (x=time, y=co2_PC, color=name)) + geom_point() + geom_line() + 
  xlab ("Year") + ylab ("CO2 Emissions Per Capita") + guides (color=guide_legend (title="Countries"))
```

For the most part for each country, either their CO2 per capita stayed constant or decreased slightly over time

### **Question 4**
```{r echo=FALSE}
gapminder %>%
  filter (!is.na (co2), !is.na (region)) %>%
  filter (time %in% c (1960,2016)) %>%
  group_by (region, time) %>%
  summarize (avgCo2_PC = mean (co2_PC))
```
Every continents average CO2 emissions increased from 1960 to 2016

### **Question 5**
```{r echo=FALSE}
gapminder %>% 
  filter (time == c (1960, 2016) & region!= "") %>% 
  group_by (region, time) %>% 
  summarise (co2_PC_avg = mean (co2_PC, na.rm=TRUE)) %>%
  ggplot (aes (x=region, y=co2_PC_avg, fill=factor(time))) + geom_bar(position="dodge", stat="identity") +
  xlab ("Continent") + ylab ("Average CO2 per capita") + guides (fill = guide_legend (title="Year"))
```

### **Question 6**
```{r echo=FALSE}
gapminder %>%
  group_by (name, region) %>%
  filter (!is.na (name), !is.na (co2_PC)) %>%
  filter (time == 2016) %>%
  mutate (avgCo2_PC = mean (co2_PC)) %>%
  summarise (n=co2_PC) %>%
  arrange (desc (n)) %>%
  head (4)
gapminder %>%
  group_by (name, region) %>%
  filter (!is.na (name), !is.na (co2_PC)) %>%
  filter (time == 2016) %>%
  mutate (avgCo2_PC = mean (co2_PC)) %>%
  summarise (n=co2_PC) %>%
  arrange (desc (n)) %>%
  tail (3)
```
The 3 largest CO2 emitters in 2019 are Qatar, Trindad and Toabago, and Kuwait. The 3 smallest CO2 emitters in 2019 are Burundi, Somalia, and the Democratic Republic of Congo

## GDP per capita
### **Question 1**
```{r echo=FALSE}
gapminder %>% 
  filter (time == 1960 & region!="") %>%
  ggplot (aes (x = lifeExpectancy, y = GDP_PC, color = region, size = totalPopulation)) + geom_point() + 
  xlab ("GDP per capita") + ylab ("Life Expectancy") + scale_size (guide = "none") +
  guides (color = guide_legend (title = "Continent")) +
  ggtitle ("GDP per capita vs. life expectancy by country in 1960")
```

For most countries besides Africa, the higher the GDP per capita is the more the Life Expectancy increases

### **Question 2**
```{r echo=FALSE}
gapminder %>% 
  filter (time == 2019 & region!="") %>%
  ggplot (aes (x = lifeExpectancy, y = GDP_PC, color=region, size=totalPopulation)) + geom_point() + 
  xlab ("GDP per capita") + ylab ("Life Expectancy") + scale_size (guide="none") +
  guides (color = guide_legend (title = "Continent")) +
  ggtitle ("GDP per capita vs. life expectancy by country in 2019")
```

### **Question 3**
Both the GDP per capita and the life expectancy of continents have increased dramatically in the last

### **Question 4**
```{r echo=FALSE}
gapminder %>% 
  filter (time == c (1960, 2019) & region!="") %>% 
  group_by (region, time) %>% 
  summarise (lifeExpectancy_avg = mean (lifeExpectancy, na.rm=TRUE)) %>%
  arrange (time)
```
The results do fit what you see in the figures

### **Question 5**
```{r echo=FALSE}
gapminder %>% 
  group_by (region) %>% 
  filter (region!="") %>%
  mutate (lifeExpectancy_prev = lag (lifeExpectancy), 
         lifeExpectancy_growth = lifeExpectancy - lifeExpectancy_prev) %>% 
  summarise (lifeExpectancy_growth_avg = mean (lifeExpectancy_growth, na.rm = TRUE))
```
The continents of the Americas and Oceania had a much higher Life Expectancy than any of the continents

### **Question 6**
```{r echo=FALSE}
gapminder %>% 
  filter (time == c(1960, 2019)) %>%
  ggplot (aes (x = GDP_PC, fill = factor (time), color=factor (time))) + 
  geom_histogram (alpha = 0.4, colour = "black",lwd = 0.75, linetype = 1,) + 
  guides (fill = guide_legend (title="Year")) + xlab ("GDP per capita")
```

### **Question 7**
```{r echo=FALSE}
gapminder %>% 
  filter (time == 1960 & name!="") %>%
  mutate (rank = rank (desc (lifeExpectancy))) %>% 
  filter (name == "United States of America") %>% 
  select (name, rank)
gapminder %>% 
  filter (time == 2019 & name!="") %>%
  mutate (rank = rank (desc (lifeExpectancy))) %>% 
  filter (name == "United States of America") %>% 
  select (name, rank)
```
The rank of the US in 1960 was 17, and the rank of the US of the US in 2019 was 46